---
title: "Splising Sites Selection"
author: "Nadzeya Boyeva"
date: "2024-11-05"
output: pdf_document
---

```{r include=FALSE}
library(dplyr)
library(edgeR)
library(Rsamtools)
library(ggplot2)
library(ggseqlogo)
library(VariantAnnotation)
library(VennDiagram)
library(grid)
source("3_SNVs_in_splicing_sites_functions.R")
#source("/media/user5/new/boyeva/r/2_splicing_sites_selection_functions.R")
```

# Sites from EEJ gene related matrices

```{r}
fd <- '/home/nadzeya/praktika'
fd116 <- '/media/user5/new/boyeva'

readTxt <- function(path) {
  read.table(file=path, 
             sep="\t",
             header=TRUE,
             quote="\"",
             as.is=TRUE)
}
```

EEJ gene related matrices contain information about exon-exon junctions which is inferred from RNA-Seq reads mapping gaps. Rows of every matrix of this kind are individual EEJs, and columns are samples. To create a set of EEJs, where splicing actually occurs in Kasumi-1 cells, we need to combine all the EEJ counts from all the samples:

```{r eval=FALSE, include=FALSE}
eej_dataset_dir <- file.path(fd116, "Dataset_of_EEJs")
all_files <- list.files(eej_dataset_dir, full.names = TRUE)
combined_df <- readTxt(all_files[[1]])

for (i in 2:length(all_files)) {
  combined_df <- full_join(combined_df, 
                           readTxt(all_files[[i]]), by = c("eej_id", 
                                                           "gene_id", 
                                                           "seqnames", 
                                                           "start", 
                                                           "end", 
                                                           "width", 
                                                           "strand"), copy=TRUE)
}

combined_df[is.na(combined_df)] <- 0

nrow(combined_df)
ncol(combined_df)
colnames(combined_df)[1:10]
```

To keep all potentially active EEJs, select low filtering threshold: an EEJ should have at least 10 reads supporting it or CPM \>= -1 (0.5 per million reads) at least in one sample.

```{r eval=FALSE, include=FALSE}
filtered_df <- filter_eej_df(combined_df,
                              use_raw=TRUE,
                              use_cpm=TRUE,
                              raw_threshold=10, 
                              cpm_threshold=-1,
                              n_samples_threshold=1)

filtered_df$mean_raw <- filtered_df$sum_raw / filtered_df$sum_samples_raw
filtered_df$mean_cpm <- filtered_df$sum_cpm / filtered_df$sum_samples_cpm

write.table(filtered_df, file = file.path(fd116, "EEJ_raw10_cpm-1_s1.txt"), sep = '\t')
str(filtered_df)
```

Repeat filtration with higher CPM threshold:

```{r eval=FALSE, include=FALSE}
filtered_df <- filter_eej_df(combined_df,
                              use_raw=TRUE,
                              use_cpm=TRUE,
                              raw_threshold=10, 
                              cpm_threshold=0,
                              n_samples_threshold=1)

filtered_df$mean_raw <- filtered_df$sum_raw / filtered_df$sum_samples_raw
filtered_df$mean_cpm <- filtered_df$sum_cpm / filtered_df$sum_samples_cpm

write.table(filtered_df, file = file.path(fd116, "EEJ_raw10_cpm0_s1.txt"), sep = '\t')
nrow(filtered_df)
```

... and with higher supporting samples number threshold:

```{r eval=FALSE, include=FALSE}
filtered_df <- filter_eej_df(combined_df,
                              use_raw=TRUE,
                              use_cpm=TRUE,
                              raw_threshold=10, 
                              cpm_threshold=0,
                              n_samples_threshold=3)

filtered_df$mean_raw <- filtered_df$sum_raw / filtered_df$sum_samples_raw
filtered_df$mean_cpm <- filtered_df$sum_cpm / filtered_df$sum_samples_cpm

write.table(filtered_df, file = file.path(fd116, "EEJ_raw10_cpm0_s3.txt"), sep = '\t')
nrow(filtered_df)
```

# Logos

To explore logos of splicing sites in selected EEJ regions, we import reference genome and extract sequences of splicing sites from it.

```{r}
ref_path = file.path(fd, "Homo_sapiens.GRCh38.dna_sm.toplevel.fa")
ref_idx_path = file.path(fd, "Homo_sapiens.GRCh38.dna_sm.toplevel.fa.fai")
file <- FaFile(ref_path, index=ref_idx_path)
fasta <- open(file)
```

We can see that GT dinucleotide in 5' SS is not that conservative with lower filtering thresholds.

```{r}
input <- "EEJ_raw10_cpm-1_s1.txt"

res <- get_SS_from_EEJ(read_from_file=TRUE, file.path(fd, input))
res$fiveSSs@seqnames <- gsub("chr", "", res$fiveSSs@seqnames)
x1 <- getSeq(x=fasta, res[[1]])
ggseqlogo(data.frame(x1), seq_type='dna', method='prob')
```

We see the same situation for 3' SS dinucleotide – normally it should be more conservative.

```{r}
res$threeSSs@seqnames <- gsub("chr", "", res$threeSSs@seqnames)
x2 <- getSeq(x=fasta, res[[2]])
ggseqlogo(data.frame(x2), seq_type = 'dna', method='prob')
```

CPM threshold higher:

```{r}
input <- "EEJ_raw10_cpm0_s1.txt"

res <- get_SS_from_EEJ(read_from_file=TRUE, file.path(fd, input))
res$fiveSSs@seqnames <- gsub("chr", "", res$fiveSSs@seqnames)
x1 <- getSeq(x=fasta, res[[1]])
ggseqlogo(data.frame(x1), seq_type='dna', method='prob')
```

```{r}
res$threeSSs@seqnames <- gsub("chr", "", res$threeSSs@seqnames)
x2 <- getSeq(x=fasta, res[[2]])
ggseqlogo(data.frame(x2), seq_type = 'dna', method='prob')
```

But if we take higher thresholds, dinucleotides are more conservative here.

```{r}
input <- "EEJ_raw10_cpm0_s3.txt"

res <- get_SS_from_EEJ(read_from_file=TRUE, file.path(fd, input))
res$fiveSSs@seqnames <- gsub("chr", "", res$fiveSSs@seqnames)
x1 <- getSeq(x=fasta, res[[1]])
ggseqlogo(data.frame(x1), method='prob')
```

```{r}
res$threeSSs@seqnames <- gsub("chr", "", res$threeSSs@seqnames)
x2 <- getSeq(x=fasta, res[[2]])
ggseqlogo(data.frame(x2), seq_type = 'dna', method='prob')
```

# Intersection of Splice Sites from Different Sources

Load EEJ data:

```{r}
eej <- readTxt(file.path(fd, "EEJ_raw10_cpm0_s3.txt"))
eej$seqnames <- gsub("chr", "", eej$seqnames)
ss_coords_eej <- get_SS_from_EEJ(read_from_file=FALSE, df=eej)
```

Load UCSC data (ucsc data):

```{r}
bed_path = file.path(fd, "ucsc.bed")
ucsc <- import(con = bed_path, format = "BED")
ucsc@seqnames <- gsub("chr", "", ucsc@seqnames)
ucsc <- ucsc[nchar(as.character(ucsc@seqnames)) < 3, ]
ucsc@seqnames <- droplevels(ucsc@seqnames)
ucsc_df <- as.data.frame(ucsc)
ucsc_df <- ucsc_df[!duplicated(ucsc_df[c("seqnames", "start", "end", "strand")]),]
ss_coords_ucsc <- get_SS_from_ucsc(read_from_file = FALSE, df=ucsc_df)
ss_coords_ucsc
```

```{r}
ucsc_df$eej_id <- paste0("chr", as.character(ucsc_df$seqnames), ":", 
                            as.character(ucsc_df$start - 1), "-", 
                            as.character(ucsc_df$end + 1), "_str", 
                            ucsc_df$strand)
```

```{r}
eej_5ss <- paste0("chr", as.character(ss_coords_eej$fiveSSs@seqnames), ":", 
                   as.character(start(ss_coords_eej$fiveSSs@ranges)), "-", 
                   as.character(end(ss_coords_eej$fiveSSs@ranges)), 
                   "_str", ss_coords_eej$fiveSSs@strand)

eej_3ss <- paste0("chr", as.character(ss_coords_eej$threeSSs@seqnames), ":", 
                   as.character(start(ss_coords_eej$threeSSs@ranges)), "-", 
                   as.character(end(ss_coords_eej$threeSSs@ranges)), 
                   "_str", ss_coords_eej$threeSSs@strand)

ucsc_5ss <- paste0("chr", as.character(ss_coords_ucsc$fiveSSs@seqnames), ":", 
                   as.character(start(ss_coords_ucsc$fiveSSs@ranges)), "-", 
                   as.character(end(ss_coords_ucsc$fiveSSs@ranges)), 
                   "_str", ss_coords_ucsc$fiveSSs@strand)

ucsc_3ss <- paste0("chr", as.character(ss_coords_ucsc$threeSSs@seqnames), ":", 
                   as.character(start(ss_coords_ucsc$threeSSs@ranges)), "-", 
                   as.character(end(ss_coords_ucsc$threeSSs@ranges)), 
                   "_str", ss_coords_ucsc$threeSSs@strand)
```

```{r}
venn.plot <- venn.diagram(
  x = list(
    eej = eej_5ss,
    ucsc = ucsc_5ss
  ),
  filename = NULL,  
  fill = c("coral", "cornflowerblue"), 
  alpha = 0.6,     
  cat.cex = 1,   
  cex = 1,
  cat.pos = c(-22, 20),
  category.names = c("RNA-Seq", "UCSC"),
  main = "5' SSs Intersection",
  main.cex = 1,
  main.pos = c(0.5, 0.98),
  main.fontfamily = "sans",
  main.fontface = "bold",
  cat.fontfamily = "sans",
  fontfamily = "sans",
)

grid.newpage() 
grid.draw(venn.plot)
```

```{r}
venn.plot <- venn.diagram(
  x = list(
    eej = eej_3ss,
    ucsc = ucsc_3ss
  ),
  filename = NULL,  
  fill = c("coral", "cornflowerblue"), 
  alpha = 0.6,     
  cat.cex = 1,   
  cex = 1,
  cat.pos = c(-22, 20),
  category.names = c("RNA-Seq", "UCSC"),
  main = "3' SSs Intersection",
  main.cex = 1,
  main.pos = c(0.5, 0.98),
  main.fontfamily = "sans",
  main.fontface = "bold",
  cat.fontfamily = "sans",
  fontfamily = "sans",
)

grid.newpage() 
grid.draw(venn.plot)
```

Here we can see which percent of SSs from RNA-Seq data and from UCSC annotation (overall, unique to each source and common) are in genes which are normally expressed in Kasumi-1 cell line.

```{r}
expressed_genes <- readTxt(file.path(fd, "RUNX1-RUNX1T1 project, list of expressed genes"))
```

```{r}
ucsc_df$refseq_id <- sapply(strsplit(ucsc_df$name, "_"), function(x) paste(x[1], x[2], sep = "_"))

unique_to_eej <- eej[eej$eej_id %in% setdiff(eej$eej_id, ucsc_df$eej_id),]
unique_to_ucsc <- ucsc_df[ucsc_df$eej_id %in%setdiff(ucsc_df$eej_id, eej$eej_id),]
common <- eej[eej$eej_id %in% intersect(eej$eej_id, ucsc_df$eej_id),]
mean(unique_to_eej$gene_id %in% expressed_genes$gene_id)
mean(unique_to_ucsc$refseq_id %in% expressed_genes$refseq_id)
mean(common$gene_id %in% expressed_genes$gene_id)
```

```{r}
mean(eej$gene_id %in% expressed_genes$gene_id)
mean(ucsc_df$refseq_id %in% expressed_genes$refseq_id)
```

```{r}
ucsc_unique <- ucsc_df[ucsc_df$eej_id %in%setdiff(ucsc_df$eej_id, eej$eej_id),]
write.table(ucsc_unique, file = file.path(fd, "ucsc_unique.txt"), sep = '\t')
```
